Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-09-06 paper/ms.tex
update ms

% 250 word max!
\begin{abstract}
  We explore a set of pre-main-sequence stars
  with sub-solar masses, rotation periods of 0.2--2 days, optical variability amplitudes
  of a few percent, and sharp light curve features that repeat for
  hundreds of rotation cycles.  The sharp features are probably caused by
  eclipsing clumps of dust or gas orbiting at the corotation radius.
  We refer to this class of star as ``complex periodic variables'' (CPVs).
  To improve our understanding of these
  enigmatic objects, we conducted a survey for CPVs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CPVs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CPVs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and for the detail with which 
  each dip's evolving amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CPVs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features could
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CPVs and of B stars
  that may have multipolar magnetic fields.
\end{abstract}

@@

\section{No significant power at 20~second cadence}

@@

\section{The CPVs are not obviously accreting}

@@

\begin{figure*}[!t]
	\begin{center}
	\subfloat{
			\includegraphics[width=0.7\textwidth]{f13a.pdf}
    }

	\vspace{-1.cm}
	\subfloat{
			\includegraphics[width=0.7\textwidth]{f13b.pdf}
    }

	\vspace{-1.cm}
	\subfloat{
			\includegraphics[width=0.7\textwidth]{f13c.pdf}
    }
	\end{center}
	\vspace{-0.4cm}
	\caption{
		{\bf Spectral age and activity diagnostics for three CPVs}.
		Wavelengths are in air;
		the continuum normalization is relative to the entire order.
		The H$\alpha$ emission strength classifies the stars as weak-lined
		T Tauris.
		The lithium detection for TIC~146539195 is consistent with its
		mass and $\beta$~Pic membership;
		the non-detections for TIC~402980664 and TIC~264599508
		are consistent with the $\approx$42\,Myr age implied by their membership
		in the Columba moving group.
	}
	\label{fig:speccutouts}
\end{figure*}

>>>>

% 250 word max!
\begin{abstract}
  We explore a set of pre-main-sequence stars
  with sub-solar masses, rotation periods of 0.2--2 days, optical variability amplitudes
  of a few percent, and sharp light curve features that repeat for
  hundreds of rotation cycles.  The sharp features are probably caused by
  eclipsing clumps of dust or gas orbiting at the corotation radius.
  We refer to this class of star as ``complex periodic variables'' (CPVs).
  To improve our understanding of these
  enigmatic objects, we conducted a survey for CPVs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CPVs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CPVs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and for the detail with which 
  each dip's evolving amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CPVs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features could
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CPVs and of B stars
  known to have multipolar magnetic fields.
\end{abstract}

@@

\section{Literature comparison}
\label{sec:litcomp}

We compared our CPV search results against previous work by
compiling a list of CPVs from both K2
\citep{2017AJ....153..152S,2018AJ....155...63S} and TESS
\citep{2019ApJ...876..127Z,2020AJ....160...86B,2021AJ....161...60S,2022AJ....163..144G,2023ApJ...945..114P}.
We counted PTFO 8-8695 as a TESS-detected CPV \citep[but see][and
references therein]{2020AJ....160...86B}.  This effort resulted in a
list of 74 unique objects that had been reported to be CPVs.  We made
no attempt to reclassify these sources based on our subjective grading
scheme.  TESS contributed 41 of the 74 objects; the remaining 33 were
from K2.

{\it How many known CPVs did we miss?}---A minority of the literature
CPVs, 19/74, were stars with 120-second TESS data acquired between
Sectors 1-55 that also met conditions~\ref{eq:one}--\ref{eq:four}.
Our blind search (Table~\ref{tab:thetable}) recovered all but two of
these 19 sources: TIC~65347864 in Tuc-Hor \citep{2023ApJ...945..114P},
and TIC~243499565 in Sco-Cen \citep{2021AJ....161...60S}.  During our
blind vetting, we manually labelled both sources as eclipsing
binaries.  Reconsidering with the knowledge of previous literature, we
believe that TIC~65347864 remains ambiguous: this source shows one
persistent broad local minimum superposed over sinusoidal spot
modulation; it could be an eclipsing binary, an RS CVn, or a CPV.

TIC~243499565 however is a bona fide CPV; our blind labeling for this
source was incorrect.  \citet{2021AJ....161...60S} made the
classification based on Sector~11 full-frame image data.  During
Sector~38, the dip phases were different, and the source
resembled an eccentric eclipsing binary (similar to TIC~193831684).  Sector~64 recently showed
the development of additional local minima in the light curve so that
it shows three dips per cycle; its CPV classification is secure.  Our
search therefore missed one star that had previously been classified
as a CPV.

{\it How many of our CPVs are new?}---Considering only our ``good''
CPVs, 35 of these 50 sources have not yet been reported in the
literature.  Considering only the 13 ``ambiguous'' CPVs, 11 are new to
the literature.  Two sources, TIC~67897871 (EPIC~202724025; RIK-90;
\citealt{2017AJ....153..152S}) and TIC~118769116 (EPIC~247343526;
\citealt{2017AJ....153..152S}), were previously only known to be CPVs
due to K2 observations.  Both periods appear to have remained
identical, within measurement precision.  The shapes have of course changed.
These two objects, combined with PTFO 8-8695, provide evidence
that CPV variability persists for at least decade timescales.

\begin{figure}[!t]
	\begin{center}
		\centering
		\includegraphics[width=0.55\textwidth]{f13.pdf}
		\vspace{-0.2cm}
		\caption{
			{\bf Brightnesses and distances of our CPVs compared
      against previous work.}  
      ``K2 literature'' corresponds to \citet{2017AJ....153..152S} and
      \citet{2018AJ....155...63S}.  ``TESS literature'' corresponds to
      \citet{2019ApJ...876..127Z,2021AJ....161...60S,2022AJ....163..144G}, and \citet{2023ApJ...945..114P}.
      ``CQVs'' and ``candidates'' are as in
      Figure~\ref{fig:catalogscatter}, though for visual clarity
      unresolved binaries are not highlighted.  PTFO 8-8695, a CPV at
      a distance of $\approx$350\,pc, is not shown.
		}
			\vspace{-0.5cm}
		\label{fig:tmagvsdist}
	\end{center}
\end{figure}

{\it How do the brightness and distance distributions
compare?}---Figure~\ref{fig:tmagvsdist} compares the TESS magnitudes
and distances for the CPVs in Table~\ref{tab:thetable} against the
previously known literature CPVs.  At the extremes, our new sources
include the closest and brightest CPVs currently known. 
All K2 CPVs are in Sco-Cen,
$\approx$100-180\,pc from the Sun.  The large gap in the TESS literature CPV
distance distribution was because the studies by
\citep{2019ApJ...876..127Z} and \citet{2023ApJ...945..114P} both
favored Tuc-Hor; many of our new CQVs are within this 40-100\,pc gap.

\section{No significant power at 20~second cadence}
\label{sec:twentysec}

@@

\section{The CPVs are not obviously accreting}
\label{sec:accretion}

@@

\begin{figure*}[!t]
	\begin{center}
	\subfloat{
			\includegraphics[width=0.7\textwidth]{f14a.pdf}
    }

	\vspace{-1.cm}
	\subfloat{
			\includegraphics[width=0.7\textwidth]{f14b.pdf}
    }

	\vspace{-1.cm}
	\subfloat{
			\includegraphics[width=0.7\textwidth]{f14c.pdf}
    }
	\end{center}
	\vspace{-0.4cm}
	\caption{
		{\bf Spectral age and activity diagnostics for three CPVs}.
		Wavelengths are in air;
		the continuum normalization is relative to the entire order.
		The H$\alpha$ emission strength classifies the stars as weak-lined
		T Tauris.
		The lithium detection for TIC~146539195 is consistent with its
		mass and $\beta$~Pic membership;
		the non-detections for TIC~402980664 and TIC~264599508
		are consistent with the $\approx$42\,Myr age implied by their membership
		in the Columba moving group.
	}
	\label{fig:speccutouts}
\end{figure*}
